---
title: "Assignment 3: Verify the Claim of Povery Change in Philadelphia"
subtitle: "Testing a published claim about neighborhood change against the data behind it"
author: "Tatar Anurakwongsri"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
library(sf)
library(ggplot2)
```

---

## Part 1 — Reconstruct

```{r}
vars <- c(total_poverty = "B17001_002",
               total_pop = "B01003_001",
               median_income = "B19013_001"
               )

data_2006 <- get_acs(
  geography = "tract",
  variables = vars,
  state = "PA",
  county = "Philadelphia",
  year = 2010,
  survey = "acs5",
  geometry = TRUE,
  output    = "wide"
)
data_2006 <- data_2006 %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

data_2013 <- get_acs(
  geography = "tract",
  variables = vars,
  state = "PA",
  county = "Philadelphia",
  year = 2017,
  survey = "acs5",
  geometry = TRUE,
  output    = "wide"
)
data_2013 <- data_2013 %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

### Notes from Article
[Article link] <https://whyy.org/articles/new-census-figures-on-philly-neighborhoods-show-inequality-high-numbers-of-whites-living-in-poverty/>

- Fairhill 61% poverty rate
- 12% --> 26% poverty rate in Eastwick 
- 2% --> 19% poverty rate in Holmesburg 
- Median household income of Tioga-Nicetown plummeted from 28,026 to 17,493

### Variables of Interest

- Poverty rate (total)
- Median household income

### Variable Codes
- total_poverty: B17001_002
- total_pop: B01003_001
- median_income: B19013_001


The article claims a 14% increase in poverty rate between the 2006-2010 and
2013-2017 ACS data in the neighborhood of Eastwick, a 17% increase in poverty rate
in Holmesburg, a drop in median household income in Tioga-Nicetown, and a 61% poverty
rate in Fairhill.

The article pointed variables of interest from the census tract: total poverty 
rate, white poverty rate, median household income, and white population. Data from
the 5-year ACS (2006-2010 & 2013-2017) and 2016 1-year ACS were used. The article
states using these data tables while referencing the 2016-2017 ACS as the then-newest
census data. I assumed some of the time period for some of the claims: Fairhill, Holmesburg,
Tioga-Nicetown and the income gap. No years were specified for these so I assumed
2013-2017 based on the title. The time periods that overlap are 2013-2017 and
2016-2017. However, the 1-year ACS data is not available at the tract level due
to the area 65,000 population requirement. The Census Bureau does not recommend
comparing different time-period lengths as they measure characteristics of the
population in different ways. 


### Load Neighborhood Data
```{r}
github_path <- "https://raw.githubusercontent.com/opendataphilly/odp-data-storage/5c4538b5f4d2a26a81b7fab2d0006d327a39e3d0/philadelphia-neighborhoods/philadelphia-neighborhoods.geojson"

neighborhoods <- st_read(github_path)
neighborhoods <- st_make_valid(neighborhoods) # Fix broken geometry of neighborhoods

neighborhoods_2006 <- data_2006 %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(st_crs(neighborhoods)) %>%
  st_join(neighborhoods) %>%
  st_drop_geometry()

neighborhoods_2013 <- data_2013 %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(st_crs(neighborhoods)) %>%
  st_join(neighborhoods) %>%
  st_drop_geometry()
```

I used a new method from the sf package called st_make_valid to fix the geometry error during data import.


### Aggregate Data by Neighborhood
```{r}
grouped_neighborhoods_2006 <- neighborhoods_2006 %>%
  group_by(LISTNAME) %>%
  summarise(
    total_pop = sum(total_popE, na.rm = TRUE),
    total_pop_moe = moe_sum(moe = total_popM, estimate = total_popE, na.rm = TRUE),
    
    total_poverty = sum(total_povertyE, na.rm = TRUE),
    total_poverty_moe = moe_sum(moe = total_povertyM, estimate = total_povertyE, na.rm = TRUE),
    
    median_income = sum(median_incomeE, na.rm = TRUE),
    median_income_moe = moe_sum(moe = median_incomeM, estimate = median_incomeE, na.rm = TRUE)
  )

grouped_neighborhoods_2013 <- neighborhoods_2013 %>%
  group_by(LISTNAME) %>%
  summarise(
    total_pop = sum(total_popE, na.rm = TRUE),
    total_pop_moe = moe_sum(moe = total_popM, estimate = total_popE, na.rm = TRUE),
    
    total_poverty = sum(total_povertyE, na.rm = TRUE),
    total_poverty_moe = moe_sum(moe = total_povertyM, estimate = total_povertyE, na.rm = TRUE),
    
    median_income = sum(median_incomeE, na.rm = TRUE),
    median_income_moe = moe_sum(moe = median_incomeM, estimate = median_incomeE, na.rm = TRUE)
  )
```

The aggregation of neighborhoods worked perfectly fine. The only data irregularities
are found in the airport area for the 5-year ACS, which yielded little to no estimates.
On the other hand, the northeast philadelphia airport reported statistics. However, 
this is likely due to the lack of residents living in the immediate proximity of the 
airport (assuming this is the international airport). Interestingly, the 2016
1-year ACS has data estimates for the airport neighborhood.

## Part 2: Test

### Calculate Poverty Rate
```{r}
grouped_neighborhoods_2006 <- grouped_neighborhoods_2006 %>%
  mutate(pov_rate = (total_poverty / total_pop) * 100,
         pov_rate_moe = moe_prop(total_poverty, total_pop, total_poverty_moe, total_pop_moe),
         pov_moe_pct = (pov_rate_moe / pov_rate) * 100) %>%
  mutate(reliability = case_when(
  pov_moe_pct < 12  ~ "High",
  pov_moe_pct < 40 ~ "Moderate",
  TRUE         ~ "Low"
  ))
  
grouped_neighborhoods_2013 <- grouped_neighborhoods_2013 %>%
  mutate(pov_rate = (total_poverty / total_pop) * 100,
         pov_rate_moe = moe_prop(total_poverty, total_pop, total_poverty_moe, total_pop_moe),
         pov_moe_pct = (pov_rate_moe / pov_rate) * 100) %>%
  mutate(reliability = case_when(
  pov_moe_pct < 12  ~ "High",
  pov_moe_pct < 40 ~ "Moderate",
  TRUE         ~ "Low"
  ))
```

### View Reliability of Each Neighborhood as a Data Table 

```{r}
library(gt)

reliability_table <- grouped_neighborhoods_2006 %>%
  mutate(
    dot = case_when(
      pov_moe_pct < 12  ~ "🟢",
      pov_moe_pct <= 40 ~ "🟡",
      TRUE              ~ "🔴"
    ),
    flag = case_when(
      pov_moe_pct < 12  ~ "Reliable",
      pov_moe_pct <= 40 ~ "Somewhat reliable",
      TRUE              ~ "Unreliable"
    ),
  ) %>%
  arrange(desc(pov_moe_pct)) %>%
  select(dot, LISTNAME, pov_rate, pov_rate_moe, pov_moe_pct, flag)

reliability_table %>%
  gt() %>%
  fmt_number(columns = c(pov_rate, pov_rate_moe), decimals = 0) %>%
  fmt_number(columns = pov_moe_pct, decimals = 1) %>%
  cols_label(
    dot               = "",
    LISTNAME          = "Neighborhood",
    pov_rate          = "Poverty Rate",
    pov_rate_moe      = "MOE",
    pov_moe_pct       = "CV (%)",
    flag              = "Reliability"
  ) %>%
  cols_align(align = "center", columns = dot) %>%
  tab_header(
    title    = "Population below 100% of the poverty level",
    subtitle = "Philadelphia Neighborhoods, ACS 2006-2010 5-year estimates"
  ) %>%
  tab_source_note(
    "Reliability thresholds follow Jurjevich et al. (2018): CV < 12% reliable, 12–40% somewhat reliable, > 40% unreliable."
  )
```

```{r}
reliability_table <- grouped_neighborhoods_2013 %>%
  mutate(
    dot = case_when(
      pov_moe_pct < 12  ~ "🟢",
      pov_moe_pct <= 40 ~ "🟡",
      TRUE              ~ "🔴"
    ),
    flag = case_when(
      pov_moe_pct < 12  ~ "Reliable",
      pov_moe_pct <= 40 ~ "Somewhat reliable",
      TRUE              ~ "Unreliable"
    ),
  ) %>%
  arrange(desc(pov_moe_pct)) %>%
  select(dot, LISTNAME, pov_rate, pov_rate_moe, pov_moe_pct, flag)

reliability_table %>%
  gt() %>%
  fmt_number(columns = c(pov_rate, pov_rate_moe), decimals = 0) %>%
  fmt_number(columns = pov_moe_pct, decimals = 1) %>%
  cols_label(
    dot               = "",
    LISTNAME          = "Neighborhood",
    pov_rate          = "Poverty Rate",
    pov_rate_moe      = "MOE",
    pov_moe_pct       = "CV (%)",
    flag              = "Reliability"
  ) %>%
  cols_align(align = "center", columns = dot) %>%
  tab_header(
    title    = "Population below 100% of the poverty level",
    subtitle = "Philadelphia Neighborhoods, ACS 2006-2010 5-year estimates"
  ) %>%
  tab_source_note(
    "Reliability thresholds follow Jurjevich et al. (2018): CV < 12% reliable, 12–40% somewhat reliable, > 40% unreliable."
  )
```


### Calculate change between two time periods and margin of error of this change

```{r}
selected_2006 <- select(grouped_neighborhoods_2006, 
                        c(LISTNAME, pov_rate_2006 = pov_rate, pov_moe_2006 = pov_rate_moe,
                          median_income_2006 = median_income, income_moe_2006 = median_income_moe))
selected_2013 <- select(grouped_neighborhoods_2013,
                        c(LISTNAME, pov_rate_2013 = pov_rate, pov_moe_2013 = pov_rate_moe,
                          median_income_2013 = median_income, income_moe_2013 = median_income_moe))

comparison_df <- inner_join(selected_2006,selected_2013, by = "LISTNAME") %>%
  na.omit(comparison_df)

pov_comparison <- comparison_df %>%
  mutate(
    pov_change = pov_rate_2013 - pov_rate_2006,
    se_1     = pov_moe_2013 / 1.645,
    se_2     = pov_moe_2006 / 1.645,
    se_diff  = sqrt(se_1^2 + se_2^2),
    moe_diff = se_diff * 1.645,
    z        = pov_change / se_diff,
    significant = abs(z) > 1.645
  )

income_comparison <- comparison_df %>%
  mutate(
    income_change = median_income_2013 - median_income_2006,
    se_1     = income_moe_2013 / 1.645,
    se_2     = income_moe_2006 / 1.645,
    se_diff  = sqrt(se_1^2 + se_2^2),
    moe_diff = se_diff * 1.645,
    z        = income_change / se_diff,
    significant = abs(z) > 1.645
  )

pov_comparison %>%
  count(significant)
```


### View Biggest Percent Changes
```{r}
pov_comparison %>%
  filter(significant) %>%
  ggplot(aes(x = reorder(LISTNAME, pov_change), y = pov_change)) +
  geom_col(aes(fill = pov_change > 0)) +
  geom_errorbar(aes(ymin = pov_change - moe_diff,
                    ymax = pov_change + moe_diff), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Statistically significant change in poverty count",
       subtitle = "2006–2010 vs 2013–2017, ACS 5-year estimates",
       x = "Neighborhoods", y = "Change in people below poverty level (%)")

```


All neighborhoods that had data were highly reliable. The only unreliable
neighborhood was the airport, which didn't have any data in the first place. Out 
of the 159 neighborhoods, all 159 had statistically significant changes to their
poverty rate. East Park, Navy Yard, Brewerytown, and Packer Park experienced the 
biggest decrease in poverty rate, while Penrose and Chinatown saw the greatest 
rise in poverty rate.

The only neighborhood that produce NAs was the airport, because its population and poverty
columns were 0 due to no one living there. Division with 0 will produce NAs as it 
is not possible. Because it was the only neighborhood that had an NA, I removed it 
from the comparison data frame. If there had been more neighborhoods with NAs, I would
explore ACS data sets from other years to find more complete data.


## Part 3: Adjudicate

### Poverty-Related Claims
```{r}
pov_comparison %>%
  filter(LISTNAME == 'Eastwick')

pov_comparison %>%
  filter(LISTNAME == 'Holmesburg')

pov_comparison %>%
  filter(LISTNAME == 'Fairhill')
```

### Income-related claims
```{r}
income_comparison %>%
  filter(LISTNAME %in% c("Tioga", "Nicetown"))
```

1. Claims that are supported

The poverty rate in Eastwick did indeed increased from 12% to 26% between the
2006-2010 and 2013-2017 ACS. The high reliability also indicates this statistic
is trustworthy. 

2. Claims that are unsupported

Fairhill did not have a 61% poverty rate in either ACS data sets, as the
poverty rate hovered around 55%. The Holmesburg claim is completely wrong. The
poverty rate was never 2% and never reach the 19% as the article suggests. Its
poverty rate was 14% in 2006-2010 and increased significantly to ~18% in
2013-2017. Lastly, the median income of Nicetown decreased significantly, but did 
not match the figures the article suggested. In 2006-2010, the median household
income of Nicetown was 99,815, decreasing to 72,443 by the time of the next
5-year ACS. Tioga had a median household income of 307,688 in 2006-2010, declining 
to 296,165 by 2013-2017, which was an insignificant decrease.

## Part 4: Rewrite

Between the 2006-2010 and 2013-2017 ACS data sets, the Eastwick neighborhood
experienced a poverty rate increase from 12% to 26%, which was statistically
significant. Holemesburg saw a 4% increase in its poverty rate, which was significant
in the neighborhood's context. Fairhill had a high, but consistent poverty rate
of 55% across the two ACS time spans. Lastly, both Tioga and Nicetown experienced
a decrease in median income: 307,688 to 296,165 for the former and 99,815 to 72,443
for the latter. Due to the high median income in Tioga, this was not a significant
change. However, the decrease in income was significant for Nicetown.

